From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-12 mechanics/paper.tex
Finish Sec 4

Finally, we include two primitive combinators for dealing with $n$-ary
products of vectors. Both of these combinators will be useful when
defining Lagrangians that are generic in the number of particles.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
    \mathrm{cross} &:& 
    \begin{array}[t]{@{}l}
      \forall m,n \mathord: \mathsf{Nat}.\\
      \quad C^\infty(X \times Y, Z) \to \\
      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~Z)
    \end{array}
  \end{array}
\end{displaymath}
The combinator $\mathrm{sum}$ sums the list of real numbers it is
given, while $\mathrm{cross}$ takes the cartesian product of two
vectors of elements of cartesian spaces, and maps the given smooth
function over them. The multiplication of natural numbers $m*n$ is
straightforwardly defined in terms of the $\mathit{natrec}$ recursion
operator. We will use $\mathrm{cross}$ in Section~\ref{sec:examples}
to define Lagrangians in terms of the interactions between systems of
particles.

@@

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G_1)G_2, 0 \rangle}

@@

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, \underline{2\pi} * Z \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cos}~e : \mathbb{R}\langle 1, 0 \rangle}
  \end{mathpar}
  FIXME: division, n-ary products (sum, zip, reduce)
  \caption{A surface syntax for smooth functions}
  \label{fig:smooth-terms}
\end{figure*}

\begin{figure}
  \centering
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \elab{z}_\Delta & = & \mathrm{const}~z \\
      \elab{x}_\Delta & = & \underline{\mathrm{project}_\Delta}(x) \\
      \elab{\mathrm{let}~\vec{x} = e_1\textrm{ in }e_2}_\Delta & = & \mathrm{pair}~\mathrm{id}~\elab{e_1}_\Delta \comp \elab{e_2}_{\Delta, \vec{x}} \\
      \elab{(e_1,...,e_n)}_\Delta &=& \underline{\mathrm{pair}_n}(\elab{e_1}_\Delta, ..., \elab{e_n}_\Delta) \\
      \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
      \elab{\vec{c}}_\Delta &=& 
    \end{array}
  \end{displaymath}
  \caption{Desugaring of the surface syntax}
  \label{fig:desugar}
\end{figure}

FIXME: define the translation

Define $\Delta$ and $\elab{\Delta}$.

Define project

\begin{theorem}
  If $\Theta |\Gamma; \Delta \vdash e : X$ then $\Theta \sepbar
  \Gamma \vdash \elab{e} : C^\infty(\elab{\Delta}, X)$.
\end{theorem}

@@

>>>>

Finally, we include two primitive combinators for dealing with $n$-ary
products of vectors. Both of these combinators will be useful when
defining Lagrangians that are generic in the number of particles.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
    \mathrm{map} &:& \forall n \mathord: \mathsf{Nat}. C^\infty(Z \times X,Y) \to C^\infty(Z \times \mathrm{vec}~n~X, \mathrm{vec}~n~Y) \\
    \mathrm{cross} &:& 
    \begin{array}[t]{@{}l}
      \forall m,n \mathord: \mathsf{Nat}.\\
      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~(X \times Y))
    \end{array}
  \end{array}
\end{displaymath}
The combinator $\mathrm{sum}$ sums the list of real numbers it is
given, while $\mathrm{cross}$ takes the cartesian product of two
vectors of elements of cartesian spaces, and $\mathrm{map}$ maps the
given smooth function over a vector. The multiplication of natural
numbers $m*n$ is straightforwardly defined in terms of the
$\mathit{natrec}$ recursion operator. We will use $\mathrm{cross}$ in
Section~\ref{sec:examples} to define Lagrangians in terms of the
interactions between systems of particles.

@@

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G_1)G_2, 0 \rangle}

@@

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, \underline{2\pi} * Z \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cos}~e : \mathbb{R}\langle 1, 0 \rangle}
  \end{mathpar}

  \medskip
  \textbf{Division, sum, and cross}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}\langle G_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \frac{e_1}{e_2} : \mathbb{R}\langle G_1G_2^{-1}, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathit{vec}~N~(\mathbb{R}\langle G, 0 \rangle)}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sum}~e : \mathbb{R}\langle G, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta, x : X \vdash e_1 : Y \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathit{vec}~N~X}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{map}~(x.~e_1)~e_2 : \mathit{vec}~N~Y}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathit{vec}~M~X \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathit{vec}~N~Y}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cross}~e_1~e_2 : \mathit{vec}~(M*N)~(X \times Y)}
  \end{mathpar}

  \caption{A surface syntax for smooth functions}
  \label{fig:smooth-terms}
\end{figure*}

\begin{figure}
  \centering
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \elab{z}_\Delta & = & \mathrm{const}~z \\
      \elab{x}_\Delta & = & \underline{\mathrm{project}_\Delta}(x) \\
      \elab{\mathrm{let}~\vec{x} = e_1\textrm{ in }e_2}_\Delta & = & \mathrm{pair}~\mathrm{id}~\elab{e_1}_\Delta \comp \elab{e_2}_{\Delta, \vec{x}} \\
      \elab{(e_1,...,e_n)}_\Delta &=& \underline{\mathrm{pair}_n}(\elab{e_1}_\Delta, ..., \elab{e_n}_\Delta) \\
      \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
      \elab{\vec{c}}_\Delta &=& \mathrm{const}~\vec{c} \\
      \elab{0}_\Delta &=& \mathrm{const}~0 \\
      \elab{e_1 + e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (+) \\
      \elab{e_1 - e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (-) \\
      \elab{e_1 e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (*) \\
      \elab{e_1 \cdot e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (\cdot) \\
      \elab{\sin~e}_\Delta &=& \elab{e}_\Delta \comp \sin \\
      \elab{\cos~e}_\Delta &=& \elab{e}_\Delta \comp \cos \\
      \elab{\mathrm{exp}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{exp} \\
      \elab{e_1 / e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp (/) \\
      \elab{\mathrm{sum}~e}_\Delta &=& \elab{e}_\Delta \comp \mathrm{sum} \\
      \elab{\mathrm{map}~(x.~e_1)~e_2}_\Delta &=& \mathrm{pair}~\mathrm{id}~\elab{e_2}_\Delta \comp (\mathrm{map}~\elab{e_1}_{\Delta,x}) \\
      \elab{\mathrm{cross}~e_1~e_2}_\Delta &=& \mathrm{pair}~\elab{e_1}_\Delta~\elab{e_2}_\Delta \comp \mathrm{cross}
    \end{array}
  \end{displaymath}
  \caption{Desugaring of the surface syntax}
  \label{fig:desugar}
\end{figure}

The combinators we have presented allow the construction of smooth
invariant functions, but doing so is painful due to the point-free
style that they force. In order to be able to actually define readable
Lagrangians in our calculus, we define a surface syntax for writing
smooth functions that is desugared into our extended version of System
F$\omega$.

We define a typing judgement of smooth terms $\Theta \sepbar \Gamma;
\Delta \vdash e : X$ by the rules in Figure~\ref{fig:smooth-terms},
where $\Theta$ and $\Gamma$ are the kinding and typing contexts of our
extension of F$\omega$, and $\Delta = x_1 : X_1, ..., x_n : X_n$ is a
context of cartesian spaces. That is, for each $x : X$ in $\Delta$, we
have $\Theta \vdash X : \mathsf{CartSp}$.

The desugaring of smooth function terms into our extension of System
F$\omega$ is via the operation $\elab{-}_\Delta$, defined in
Figure~\ref{fig:desugar}. In this definition, we make use of two
derived combinators for projection and pairing:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \underline{\mathrm{project}_{x : X, \Delta}}(x) & = & \mathrm{proj_1} \\
    \underline{\mathrm{project}_{y : Y, \Delta}}(x) & = & \mathrm{proj_2} \comp \underline{\mathrm{project}_\Delta}(x) \quad (x \not= y)
  \end{array}
\end{displaymath}
and
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \underline{\mathrm{pair}_1}(e) & = & e \\
    \underline{\mathrm{pair}_{n+1}}(e, \vec{e}) & = & \mathrm{pair}~e~(\underline{\mathrm{pair}_n}(\vec{e}))
  \end{array}
\end{displaymath}
The following theorem follows easily by induction on the typing
derivations of smooth terms, where $\elab{\Delta} = \elab{x_1 : X_1,
  ..., x_n : X_n} = X_1 \times ... \times X_n$:
\begin{theorem}
  If $\Theta |\Gamma; \Delta \vdash e : X$ then $\Theta |
  \Gamma \vdash \elab{e}_\Delta : C^\infty(\elab{\Delta}, X)$.
\end{theorem}

@@

Quantum programming languages

Information effects
